---
title: "MOCA/DAWA cluster randomized trial"
author: "A.Amstutz"
format:
  html:
    toc: true
    toc-float: true
    code-fold: true
    keep-md: true
  pdf:
    toc: true
editor: visual
---

## Packages

```{r}
#| message: false
#| warning: false
RNGkind("L'Ecuyer-CMRG") # simstudy
set.seed(19287) # for reproducibility
library(simstudy)
library(parallel) # for parallelization of core (max 8 on my laptop)

library(lme4)
library(marginaleffects) # marginal standardization
library(insight) # robust SE
library(geepack) # to tackle alternative and more robust (but less efficient?) estimands

library(dplyr)
library(pwr)
library(ggplot2)
library(kableExtra)
```

## DAWA cluster randomized trial (CRT)

Interventions on the level of health care workers to reduce antibiotic prescriptions at health facilities in Zanzibar. Multi-arm with 2 interventions:

-   **Control**: Standard of care

-   **Intervention 1**: eHealth tool (CDSS & nudging)

-   **Intervention 2**: eHealth tool (CDSS & nudging) + AMR stewardship clubs

### Parameters

-   Eligible participants: Patients attending the dispensary with acute infectious illness

-   Power for subgroup of kids under 5 years (special subgroup of interest), ca. 33% of all attending patients

-   Cluster size of eligible overall participants: 80-500 per cluster per month (cluster size variation!)

-   Cluster size of eligible kids under 5 years: 26-165 per cluster per month

-   Max. 39 clusters (health dispensaries) due to feasibility/budget

-   Binary outcome: Proportion of patients prescribed an antibiotic at first presentation to care

-   Baseline prescription rate in control clusters: 75%, based on data from existing facilities but some variability

-   Expected delta Control to Intervention 1: 25 percentage points, based on previous studies in same setting

-   Expected delta Control to Intervention 2: 30 percentage points

-   Intervention 1 vs Intervention 2 not of primary interest

-   Desired power min. 80%

-   ICC for AB prescription: 0.2, based on previous studies in same setting, but some uncertainty

-   Mean cluster size: 40/month, but high variability (ratio of standard deviation of cluster sizes to mean of cluster sizes: 0.5-0.6)

-   We expect intervention effect to manifest 3-4 months after baseline

### Design considerations

-   3-arm vs 2x 2-arm?

<!-- -->

-   Recruitment bias?

-   Secular trend?

-   Which pair-wise comparisons to power for?

-   Multiplicity?

```{r}

```
